---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setoptions, echo=FALSE}
library(knitr)
opts_chunk$set(echo=TRUE)
```

## Loading and preprocessing the data
```{r readdata}
amdata <- read.csv("activity.csv")
library(dplyr, warn.conflicts=FALSE)
#gp_amdata <- group_by(amdata, date)
#tamdata <- transform(amdata, datef=factor(date)) #factor/group by date
```

## What is total number of steps taken per day?
```{r sumsteps}
#sumdata <- summarize(gp_amdata, sumsteps=sum(steps))
sumdata <- aggregate(amdata$steps, by=list(amdata$date), sum)
names(sumdata) <- c("date", "sumsteps")
sumdata
```

## Histogram for steps
```{r histogram}
hist(sumdata$sumsteps, breaks=nrow(sumdata), xlab="Sum of Steps", main="Histogram for sum of steps")
```

## What is mean steps taken per day?
```{r meansteps}
meandata <- aggregate(amdata$steps, by=list(amdata$date), mean)
names(meandata) <- c("date", "meansteps")
meandata
```

## What is median steps taken per day?
```{r mediansteps}
mediandata <- aggregate(amdata$steps, by=list(amdata$date), median)
names(mediandata) <- c("date", "mediansteps")
mediandata
```

## What is the average daily activity pattern?
```{r daily-activity}
apdata <- aggregate(amdata$steps, by=list(amdata$interval), mean, na.rm=TRUE)
names(apdata) <- c("interval", "AverageSteps")
#apdata
plot(apdata$interval, apdata$AverageSteps, type="l", xlab="Interval", ylab="AverageSteps", main="Average Daily Activity Pattern")
apdata[which.max(apdata$AverageSteps),]
```


## Imputing missing values
**Strategy for flling missing values** - *Replace all NAs with mean of the related 5 minute interval*
```{r imputing-missing-values}
#no. of NA rows
nrow(amdata[is.na(amdata$steps),])
#calculate mean steps by interval
apdata <- aggregate(amdata$steps, by=list(amdata$interval), mean, na.rm=TRUE)
#merge original data with mean steps by interval data
new_amdata <- merge(amdata, apdata, by.x="interval", by.y="Group.1")
#replace NA values with mean of steps for that interval
new_amdata[is.na(new_amdata$steps),] <- mutate(new_amdata[is.na(new_amdata$steps),], steps=x)
#sum steps for each day
sumdata <- aggregate(new_amdata$steps, by=list(new_amdata$date), sum)
names(sumdata) <- c("date", "sumsteps")
hist(sumdata$sumsteps, breaks=nrow(sumdata), xlab="Sum of Steps", main="Histogram for sum of steps")
```

New mean of steps taken per day
```{r new-meansteps}
meandata <- aggregate(new_amdata$steps, by=list(new_amdata$date), mean)
names(meandata) <- c("date", "meansteps")
meandata
```

New median of steps taken per day
```{r new-mediansteps}
mediandata <- aggregate(new_amdata$steps, by=list(new_amdata$date), median)
names(mediandata) <- c("date", "mediansteps")
mediandata
```

The original and new values for mean and median differ a bit as NAs are placed with means.

## Are there differences in activity patterns between weekdays and weekends?
```{r weekdays-n-weekend-patterns}
#convert date to weekday and append as column
weekdayscol <- data.frame(weekdays(as.Date(amdata$date)))
new_amdata <- cbind(amdata, days=weekdayscol)
#rename weekday column
new_amdata <- rename(new_amdata, days=weekdays.as.Date.amdata.date..)
new_amdata$days <- as.character(new_amdata$days)
new_amdata[new_amdata$days %in% c("Saturday", "Sunday"),]$days <- "weekend"
new_amdata[new_amdata$days %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday"),]$days <- "weekday"
new_amdata <- transform(new_amdata, days=factor(days))

apdata <- aggregate(new_amdata$steps, by=list(new_amdata$interval), mean, na.rm=TRUE)
names(apdata) <- c("interval", "AverageSteps")

all_amdata <- merge(new_amdata, apdata, by.x="interval", by.y="interval")
library(lattice)
xyplot(AverageSteps ~ interval | days, data=all_amdata, type="l", layout=c(1,2), xlab="Interval", ylab="Number of steps")


```
